---
title: "week3"
author: "Shengtong"
date: "October 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## review 

* Review about functions (nesting)
```{r, echo=T}
myFirstFun <- function(n) {
  return(n*n)
}
mySecFun <- function(v) {
  u <- myFirstFun(v)
  return(u)
}
v <- c(1,2,3,4,5)
mySecFun(v)
```

#####################################################
*  Loops in R
######################################################

 Rule of thumb: if you need to replicate something > 3 times
 a loop is faster, more compact, more readable, more maintainable,
 and saves you some typing!

### Example 1

*  This "initializes" the object "w" so that R knows what w "stands for."
```{r, echo=T}
w <- NULL
#### This is called a "for" loop
for(i in 1:10){
  w[i] <- i+10
}

for(i in 1:10){
  v[i] <- i+10
}
```
*  What do you think w looks like?
```{r, echo=T}
w <- NULL
for(hello in 1:10){
  w[hello] <- hello
}
w
```
 Now what does w look like?

### Example 2:

*  You can embed loops as well.
```{r, echo=T}
w <- NULL
counter <- 1
for(j in 11:20){
  for(i in 1:10){
    cat("i=", i, "and j=", j, "\n")
    w[counter] <- i + j
    counter <- counter + 1
  }
}
```

*  Think through these two loops carefully.  What is the difference between this loop and the next loop?

```{r, echo=T}
w <- NULL
counter <- 1
for(j in 11:20){
  for(i in 1:10){
    w[counter] <- i + j
  }
  counter <- counter + 1
}
w
```

*  It is usually a better idea to pre-allocate a variable of the appropriate size. R has to reallocate memory at each iteration
```{r, echo=T}
w <- rep(NA, length(1:10))
for(i in 1:10){
  w[i] <- i+10
}
w
```


```{r, echo=T}
system.time({
  w <- c()
  for(i in 1:10^6)
    w[i] <- i+10
  }
)

system.time({
  w <- rep(NA, 10^6)
  for(i in 1:10^6)
    w[i] <- i+10
  }
)

```

## In class example: 

* Construct a 10 x 10 multiplication table  using a nested loop.

```{r, echo=T}
my_mat <- matrix(NA, nrow=10, ncol=10)
for(rows in 1:10) {
  for(cols in 1:10) {
    my_mat[rows, cols] <- rows * cols
  }
}
my_mat
```

## Example 3: 

 while loop: Based on onset and verification of a logical 
 ;condition that is tested at beginning or end of the loop
 ;Make sure the terminiation is explicitly set by testing a condition!

*  An example of a "while" loop

```{r, echo=T}
w <- 100
z <- 5
while(w > 20){
  print(w)
  w.plus.z <- w + z
  w <- w - 1
}
```

*  Make sure that your loop can be "closed." The following loop will run forever. DON'T actually run it!!!!!

```{r, echo=T, eval=F}
w <- 100
z <- 5
while(w > 20){
w.plus.z <- w + z
w <- w + 1
}

```

## Example 4:

*  using break or next
```{r, echo=T}
m=20
for(k in 1:m) {
  if(!k %% 2)
    next
  print(k)
}
```

```{r, echo=T}
set.seed(12345) ## Set seed for random number generator
m <- rnorm(10)
for(i in 1:length(m)) {
  if(m[i] < 0)
    break
  print(m[i])
}
```

## In class example: 

Construct a lower triangular matrix with a  10 x 10 multiplication table using a nested loop.

```{r, echo=T}
my_mat <- matrix(NA, nrow=10, ncol=10)
for(rows in 1:10) {
  for(cols in 1:10) {
    if(rows >= cols) next
    my_mat[rows, cols] <- rows * cols
  }
}
my_mat
my_mat[1,3]
```


* There is also the replicate function for repeated evaulation of the same thing:
```{r, echo=T}
m <- 10
n <- 5
mymat <- replicate(m, rnorm(n))
```

*  This is equivalent!
```{r, echo=T}
mymat <- matrix(NA, nrow=5, ncol=10)
for(i in 1:ncol(mymat)) {
  mymat[,i] <- rnorm(5)
}
```

* This is also equivalent!
```{r, echo=T}
mymat <- matrix(NA, nrow=5, ncol=10)
for(cols in 1:ncol(mymat)) {
  for(rows in 1:nrow(mymat)) {
    mymat[rows,cols] <- rnorm(1)
  }
}
```

* Checking on vector interpretaition of matrices
```{r, echo=T}
my_mat <- matrix(1:20, nrow=4, ncol=5)
for(rows in 1:4) {
    my_mat[rows, ] <- rnorm(5)
}

my_mat <- matrix(1:20, nrow=4, ncol=5)
for(rows in 1:4) {
  my_mat[rows] <- rnorm(5)
}


for(index in 1:20) {
  my_mat[index] <- my_mat[index] * 10
}
```


## Example 5:

###  Vectorization

In R, everything is a vector! Many loop constructs can be made 
 implicit by using vectorization 

*  Consider this:

```{r, echo=T}
v1 <- runif(10)
v2 <- runif(10)
v3 <- c()
for (i in 1:10) {
  v3[i] <-v1[i] + v2[i]
}
v3
## versus this:
v3 = v1 + v2
v3
```

*  Consider this

```{r, echo=T}
A <- matrix(1:20, nrow=5, ncol=4)
B <- matrix(21:40, nrow=5, ncol=4)
matsum <- matrix(0, nrow=5, ncol=4)
for(i in 1:5) {
  for(j in 1:4) {
    matsum[i,j] <- A[i,j] + B[i,j]
  }
}
matsum
## versus this:
matsum <- A + B
matsum
```

 Why is vectorization faster? R is an interpreted language!

 Let's consider an example where we had some artificial "signal" to
 random noise

*  This is a bad loop with 'growing' data
```{r, echo=T}
set.seed(42)
m <- 10
n <- 10
# Create matrix of normal random numbers
mymat <- replicate(m, rnorm(n))
# Transform into data frame
mydframe <- data.frame(mymat)
for (i in 1:m) {
  for (j in 1:n) {
    mydframe[i,j]<-mydframe[i,j] + 10*sin(0.75*pi)
  }
}
# 
```

```{r, echo=T, eval=F}
# ## Here is the vectorized form:
mydframe_vec <- data.frame(mymat)
mydframe_vec <- mydframe_vec + 10*sin(0.75*pi)
mydframe_vec
```

```{r, echo=T, eval=F}
## What happens when you add a matrix and a vector?
m <- matrix(1:20, nrow=4, ncol=5)
v <- c(1,10,100,1000,10000)
m + v # v is treated as a matrix 
```

## In class example  

 Quantify the execution time for the loop and vectorized
 versions of the above example using system.time.
 Try varying m and n (m=n=10, 100, 1000). 



* This is a bad loop with 'growing' data
```{r, echo=T, eval=F}
time_check <- function(m,n) {
  set.seed(42)
  # Create matrix of normal random numbers
  mymat <- replicate(m, rnorm(n))
  # Transform into data frame
  mydframe <- data.frame(mymat)
  time_1 <- system.time({
    for (i in 1:m) {
      for (j in 1:n) {
        mydframe[i,j]<-mydframe[i,j] + 10*sin(0.75*pi)
      }
    }
  })
  time_2 <- system.time({
    mydframe_vec <- mydframe + 10*sin(0.75*pi)
  })
  return(list(loop = time_1, vectorized=time_2))
}

time_check(10,10)
time_check(100,100)
time_check(1000,1000)
```


## In class example  (from the Euler Project)

If we list all the natural numbers below 10 that are multiples  of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Find the sum of all the multiples of 3 or 5 below 1000.

* First write this in a loop.
* Second, write this in vectorized form.

* Loop
```{r, echo=T, eval=F}
keep <- c()
counter <- 1
while(counter < 1000) {
  if(counter %% 3 & counter %% 5) {
    counter <- counter + 1
  } else {
    keep <- c(keep, counter)
    counter <- counter + 1
  }
}
sum(keep)
```


## Time permitting, Chapter 7: Logic and control

*  Simple logical operators
```{r, echo=T, eval=F}
a = TRUE
 b = FALSE
# 
 a & b
 a | b
 !a & b
 !a | !b
```


*  Other operators that evaluate to TRUE or FALSE
```{r, echo=T}
 x <- 10
 x < 20
# 
 x <- 1:10
 x < 5
 x == 6
# 
 #### if-else conditions
 x <- 5
 ifelse(x==4, 1, 0)
 
 x <- 1:10
 ifelse(x<=4, 1, 0)
# 
 y <- NULL
 x <- seq(0.1, 1, by=0.1)
 for(i in 1:10){
 	if(x[i] < 0.5){
 		y[i] <- x[i]^2	
 	}else{
 		y[i] <- sqrt(x[i])}
 }
#}
 ```