---
title: "Week 1"
author: "Shengtong"
date: "September 17, 2018"
output:
  pdf_document: default
#  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R tutorial 

Run through examples in chap 1; practice R as a calculator

```{r, echo=T}
3+4
6^3
8/2
(2+5)*3
ab=3
cd=8
ab+cd

```

In R, lowercase and uppercase characters are different 

```{r, echo=T}
ab=3
AB=30
cd=8
AB+cd
```
"=" means the calculating the value on the right and storing it to the name on the left, or called "assigment statement".  

```{r, echo=T}
num=6
num=num+1 # it doesn't make sense in math 
num
```

## vectors 

* define a vector and display its characteristics 

```{r, echo=T}

my_vector <- c(1,3,5,8,10,-5,38623) # c() command says "combine" values in paratheses; do not use commas in large numbers and it's used to separate different numbers   
length(my_vector)
sum(my_vector)

```

*  Set random seed and generate some random normal draws

```{r, echo=T}
set.seed(13579)
my_normal_vector <- rnorm(n=length(my_vector))
my_normal_vector
```


*  Vector multiplication (elementwise)

```{r, echo=T}

my_vector * my_normal_vector # multiplication operation applies to very element 

```

* Inner product

```{r, echo=T}
my_vector %*% my_normal_vector
```


In R console, when a line is too long, you may continue it on the next line by typing Enter key at a place where the command is obviously not complete. 


## matrix 

*  Create matrix: 5 x 4 by  Normal distribution: mean=10, sd=4
```{r, echo=T}
my_matrix <- matrix(rnorm(n=5*4, mean=10, sd=4),
                    nrow=5, ncol=4)
```

* Create another matrix: 5 x 4 by  rexp distribution
```{r, echo=T}
gen_exp <- rexp(n=5*4, rate=1)
my_matrix_2 <- matrix(gen_exp, ncol=4, nrow=5)
```

*  Double check the product functions (elementwise multiplication)
```{r, echo=T}
my_matrix
my_matrix_2
my_matrix * my_matrix_2
```



*  Make use of head and tail commands
```{r, echo=T}
my_huge_matrix <- matrix(rgamma(10000,
                                shape=1), 
                         ncol=10)
head(my_huge_matrix)
tail(my_huge_matrix)

```

* sequence of values 

```{r, echo=T}
## Make sequence of values
1:10
0:10
10:0
-5:5
seq(from=-5, to=50, by=5)
seq(from=-5, to=50, length=50)

```

* data frame vesus matrix 

```{r, echo=T}

## Data frame versus matrix
my_df <- data.frame(my_matrix, letters=c("A", "b", "c", "d","e"))

## Character vectors
my_name_vector <- c("Anne", "Bob", "Charles")

## Check the class of objects
class(my_name_vector)
class(my_normal_vector)
class(my_huge_matrix)
class(my_df)

```


## plots 

* Basic plotting

```{r, echo=T}

#-------------------------------------------------------------
## Basic plotting

## Histogram: hist
## ?hist
## Histogram of first five columns of my_huge_matrix

hist(my_huge_matrix[,c(1:5)])
hist(my_huge_matrix[,1])
hist(my_huge_matrix[,2])
hist(my_huge_matrix[,3], main="Better histogram",
     xlab="Value", breaks=50, 
     col="cornflowerblue")

par(mfrow=c(2,3))
hist(my_huge_matrix[,1])
hist(my_huge_matrix[,2])
hist(my_huge_matrix[,3], main="Better histogram",
     xlab="Value", breaks=50, 
     col="cornflowerblue")

```

* Scatterplots 

```{r, echo=T}

## Scatterplots
data(iris)
?plot
## Plot petal length vs petal width
col_vector <- rep(NA, nrow(iris))
setosa_index <- which(iris$Species == "setosa")
versicolor_index <- which(iris$Species == "versicolor")
virginica_index <- which(iris$Species == "virginica")
col_vector[setosa_index] <- "blue"
col_vector[versicolor_index] <- "red"
col_vector[virginica_index] <- "black"

plot(x=iris$Petal.Length, 
     y=iris$Petal.Width,
     col=col_vector, pch=19)
legend("topleft", pch=19,
       col=c("blue","red","black"),
       legend=c("Setosa", "Versicolor", "Virginica"))

```

```{r, echo=T}

## Plot Petal length vs width, coloring by sepal length
col_vector <- rep(NA, nrow(iris))
high_index <- which(iris$Sepal.Length > 
                      mean(iris$Sepal.Length))
col_vector[high_index] <- "red"
col_vector[-high_index] <- "blue"
shape_vector <- rep(NA, nrow(iris))
shape_vector[setosa_index] <- "S"
shape_vector[versicolor_index] <- "E"
shape_vector[virginica_index] <- "I"

plot(x=iris$Petal.Length, 
     y=iris$Petal.Width,
     col=col_vector, pch=shape_vector)
legend("topleft", pch=19,
       col=c("blue","red"),
       legend=c("Low sepal length", 
                "High sepal length"))

plot(x=1:nrow(iris), y=sort(iris$Petal.Width),
     type="l")

```

* Boxplots 

```{r, echo=T}

## Boxplots
boxplot(iris$Petal.Length ~ iris$Species,
        col=c("red", "blue", "green"))

## Density plots
## Hint: which function to select species
## lines()
plot(density(iris$Petal.Length[which(iris$Species == 
                                       "setosa")]),
     xlim=c(0,8))
lines(density(iris$Petal.Length[which(iris$Species == 
                                       "virginica")]))
lines(density(iris$Petal.Length[which(iris$Species == 
                                       "versicolor")]))

```

