---
title: "lecture11"
author: "Shengtong"
date: "2018-11-12"
output: workflowr::wflow_html
---

```{r, echo=T, eval=F}
########################################################
#### Read https://deanattali.com/blog/building-shiny-apps-tutorial/
####
#### Still working on projects
####
#### HW7 (last one!) due next week
######################################################


#=============================================================
## Interactivity in R: In-class slides
#=============================================================

#-------------------------------------------------------------
## In class:
## 1. Create several ggplots of the iris data (scatterplot,
##   histogram, boxplot, ...) 
## 2. Experiment with changing the hover-over information with tooltip
#-------------------------------------------------------------

library(tidyverse)
library(plotly)

#=============================================================
## Shiny:
##
## R package from RStudio that is a web application framework for R
## R code -> interactive web page (no HTML/CSS/Javascript knowledge required)
##
## Let's look at http://www.showmeshiny.com/ for some ideas
#=============================================================

library(shiny)
runExample("01_hello")

## Shiny apps are composed of two parts: 
##   (1) web page that shows app to user = UI (user interface)
##   (2) computer that powers the app (either your own or on a server somehwere else) = server

## Tutorial we'll follow in class:
## https://deanattali.com/blog/building-shiny-apps-tutorial/
```

## Create an empty Shiny app

```{r, echo=T}
library(shiny)
ui <- fluidPage()
server <- function(input, output) {}
shinyApp(ui = ui, server = server)
```


## Load the dataset

```{r, echo=T}
bcl <- read.csv("shiny//bcl-data.csv", stringsAsFactors = FALSE)
print(str(bcl))
```

## Lots of other helpful stuff:
##
## Layout guides: http://shiny.rstudio.com/articles/layout-guide.html
## http://rmarkdown.rstudio.com/authoring_shiny.html
## http://shiny.rstudio.com/images/shiny-cheatsheet.pdf
## http://shiny.rstudio.com/tutorial
## http://rstudio.github.io/shinythemes/

## Ideas for cool add-ons: 
##  -- DT::dataTableOutput() for interactive table
##  -- leaflet for interactive maps
##  -- flexdashboard for interactive dashboards
##  -- ggplotly for interactive ggplot2 graphics

## Two most common "why isn't my app running?!" problems:
## (1) comma-separate all the elements
## (2) do not add comma to the last element


#=============================================================
## Next time: 
## Machine learning in R, clustering (K-means, hierarchical clustering),
##   data mining
#=============================================================

```
