---
title: "week2"
author: "Shengtong"
date: "September 26, 2018"
output:
  pdf_document: default
#  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## working directory 

```{r, echo=T, eval=F}

#------------------------------------------------------------------
## Working directories in R
#------------------------------------------------------------------

### What working directory are we in?
# getwd() ## Get working directory
# dir()   ## What else is in that directory?

## Set working directory to a new place
# !!! You need to change this!
# setwd('/Users/raua/Desktop/PH718_AR/Lecture2')
```

## R script 

* a prepared list of R commands that are processed sequentially by R from top to bottom and they can be typed, edited and saved as a file. 

* only exist for windows and mac, not in Unix or Linux operating systems.




* Save R-script as "Lecture2_Rcode.R" to PH718 directory

File -> save as > folder 

```{r, echo=T}
#### Enter the following lines
x <- 1:10
y <- 1:10
plot(x,y)

### Re-save this script

### Now run the source()command as below, without the ## in front.
##source("Lecture2_Rcode.R)

```

* two ways to run R code in Unix/Linux 

(1) copy the script and paste it at console prompt 
(2) save the script in a folder and use the command source()

```{r, echo=T, eval=F}
source("path-to-the-folder/**.R")
```


```{r, echo=T}

### Now try running, without the comment ### bars in front.
### Why do we get a "node stack overflow" error?

### Make a new folder in your PH718 directory called More_R_Commands.

### Create a new R script in that folder called "plot.R"
### Use the same 3 lines as above
### x <- 1:10
### y <- 1:10
### plot(x,y)

### Now run 
### source("More_R_Commands/plot.R")

### Try the following commands
### objects()
### ls()
### rm(list=c("x", "y"))
### ls()

```


## Functions in R 


* sum function

```{r, echo=T}
#------------------------------------------------------------------
## Functions in R
#------------------------------------------------------------------


### Here is a simple example
example.sum <- function(a, b){
return(a + b)
}

x <- 1:10
y <- 11:20
example.sum(x,y)
```



```{r, echo=T}
### Naming conventions. Why not write example.sum() as:
example.sum <- function(x, y, z){
  tmp <- x + y
  tmp2 <- tmp + z
  return(tmp2)
}
x <- 1:10
y <- 11:20
z <- 100
example.sum(x,y,z)
```


```{r, echo=T, eval=F}
### Try this

example.diff.sum <- function(x, y){
  newy = example.sum(x,y) - 10
  return(newy)
}
example.diff.sum(x, y)
```

```{r, echo=T}

### This function example.diff.sum() works, but it is very BAD form.
### Variable names can get complicated VERY quickly, if you're not
### conscientious about it. The following would be much easier to 
### understand. Note: when possible, try to name your variables 
### and functions informatively and in a manner that they are easy to track.

example.sum = function(a, b){
return(a + b)
}

example.diff.sum = function(c, d){
e = example.sum(c, d) - 10
return(e)
}

x <- 1:10
y <- 11:20
example.diff.sum(x, y)
```


```{r, echo=T}

## Example together: Hardy-Weinberg problem #3.6 from book

HWE <- function(p) {
  stopifnot(is.numeric(p))
  stopifnot(p >= 0)
  stopifnot(p <= 1)
  prob_AA <- p^2
  prob_AB <- 2*p*(1-p)
  prob_BB <- (1-p)^2
  return(c(prob_AA, prob_AB, prob_BB))
}
HWE(0)
HWE(0.5)
#HWE(-5)
#HWE("andrea")
```


```{r, echo=T}

## Write a function to convert Celsuis to Fahrenheit and vice versa

C_to_F <- function(temp) {
  return((temp * 1.8) + 32)
}
  
F_to_C<- function(temp) {
  return((temp - 32)/1.8)
}

par(mfrow=c(1,2))
plot(1:100, C_to_F(1:100))
plot(-40:212, F_to_C(-40:212))
```

```{r, echo=T}

## type_of_conversion can be "F_to_C" or "C_to_F"
temp_conversion <- function(temp, type_of_conversion="F_to_C") {
  if(!type_of_conversion %in% c("F_to_C", "C_to_F")) {
    stop("STOP!!! I only to C to F or F to C.")
  }
  if(type_of_conversion == "F_to_C") {
    new_temp <- F_to_C(temp)
  }
  if(type_of_conversion == "C_to_F") {
    new_temp <- C_to_F(temp)
  }
  return(list(temp=new_temp, conversion=type_of_conversion))
}
temp_conversion(100, type="C_to_F")
```


## loops in R 


```{r, echo=T}

#------------------------------------------------------------------
## Loops in R
#------------------------------------------------------------------

#### An introductory example

### This "initializes" the object "w" so that R knows what w "stands for."

w <- NULL

#### This is called a "for" loop

for(i in 1:10){
	w[i] <- i+10
}

### What do you think w looks like?
w 
```

```{r, echo=T}
w <- NULL
for(hello in 1:10){
w[hello] <- hello
}
## Now what does w look like?
w
```
*  You can embed loops as well.

```{r, echo=T}
w <- NULL
counter <- 1
for(j in 11:20){
  for(i in 1:10){
    w[counter] <- i + j
    counter <- counter + 1
  }
}

#### Think through these two loops carefully. 
#### What is the difference between this loop and the last loop?
w
```

```{r, echo=T}
w <- NULL
counter <- 1
for(j in 11:20){
	for(i in 1:10){
	w[counter] <- i + j
}
	counter <- counter + 1
}
w
```

*  An example of a "while" loop

```{r, echo=T}
w <- 100
z <- 5

while(w > 20){
	w.plus.z <- w + z
	w <- w - 1
}
w
```

```{r, echo=T, eval=F}

### Make sure that your loop can be "closed."
### The following loop willrun forever. DON'T actually run it!!!!!
#w <- 100
#z <- 5
#while(w > 20){
#	w.plus.z <- w + z
#	w <- w + 1
#}
```


*  If we have time, let????Ts talk about indices and the which() function

```{r, echo=T}
 x <- rnorm(100)
 x[1:10]
 x[20:30]
 index <- which(x < 0)
 x[index]

```
