---
title: "lecture4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(devtools)
library(edgeR)

```


Read Chapters 3 and 5: http://r4ds.had.co.nz/transform.html


##  review about vectorization

 In class example  (from the Euler Project)
 If we list all the natural numbers below 10 that are multiples 
 of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23.
 Find the sum of all the multiples of 3 or 5 below 1000.
 First write this in a loop.

*  Loop form
```{r, echo=T}

natural_numbers <- 1:1000
sum_value <- 0
for(i in 1:length(natural_numbers)) {
  ## Is natural numbers divisible by 3?
  ## Is natural numbers divisible by 5?
  if(!natural_numbers[i] %% 3 | !natural_numbers[i] %% 5) {
    print(natural_numbers[i])
    sum_value <- sum_value + natural_numbers[i]
  }
}
sum_value
```

```{r, echo=T}
## Vectorized form
new_vector <- sum(ifelse(!natural_numbers %% 3 | !natural_numbers %% 5,
                     natural_numbers, 0))

new_indices <- ifelse(!natural_numbers %% 3 | !natural_numbers %% 5,
       TRUE, FALSE)
natural_numbers[which(new_indices == TRUE)]
```

# Logic and control

```{r, echo=T}

##########################################################
#### Chapter 7: Logic and control
##########################################################

# #### Simple logical operators
a = TRUE
b = FALSE

a & b
a | b
!(a & b)
!a | !b

# #### Other operators that evaluate to TRUE or FALSE

x <- 10
x < 20

x <- 1:10
x < 5
x == 6

# #### if-else conditions
x <- 5
ifelse(x==4, 1, 0)

x <- 1:10
ifelse(x<=4, 1, 0)

y <- NULL
x <- seq(0.1, 1, by=0.1)
for(i in 1:10){
	if(x[i] < 0.5){
		y[i] <- x[i]^2
	}else{
		y[i] <- sqrt(x[i])}
}

# What does this evaluate to, and why?
(TRUE + TRUE) * FALSE
```

## Install R Packages 

```{r, echo=T, eval=F}

##########################################################
#### Installing packages in R
##########################################################

## Package = fundamental unit of shareable code
## Bundles together code, data, documentation, tests, and is easy to share with others

## 3 main repositories for packages in R:
## CRAN: https://www.r-project.org/
## Bioconductor: http://bioconductor.org/ (packages oriented towards analysis of biological data)
## GitHub: at indivdual user's accounts, for example https://github.com/andreamrau/coseq
## Also there is now Neuroconductor: https://neuroconductor.org/ for imaging analysis

## There are > 6000 packages at CRAN alone! And growing rapidly!!
## => Chances are, someone has probably already solved a problem you're working on
## Make sure to give credit to package developers if you use their work! (citations etc)

## Let's practice installing a couple of packages:

# ## Tidyverse and devtools package from CRAN
install.packages(c("tidyverse", "devtools"))  ## You only need to do this once per package
library(tidyverse)
library(devtools)
?tidyverse

## edgeR RNA-seq expression analysis package from Bioconductor
## This is just an example, we won't actually be using edgeR today
source("http://www.bioconductor.org/biocLite.R")   ## You only need to do this once per package
biocLite("edgeR")  ## You only need to do this once per package
library(edgeR)
vignette("edgeR")

# ## Install a GitHub package (using the devtools package above)
# ## This is just an example, we won't actually be using HTSFilter today
install_github("andreamrau/HTSFilter")  ## You only need to do this once per package
library(HTSFilter)
```

```{r, echo=T}


##########################################################
####  Core structures for analysis: data.frames and tibbles
##########################################################

## First of all: What is the tidyverse? https://www.tidyverse.org/
##    - A "meta-package" that is an optionated collection of several R packages for data science
##    - Includes ggplot2 (plotting), dplyr (transformating/summarizing dataframe content), tidyr (tidying dataframes), purr, tibble
##    - Founded by Hadley Wickham and many others at RStudio

# ## Pipe operator: f(x) can be rewritten as x %>% f 
data(iris)
head(iris)
iris %>% head()

mean.sepal <- mean(iris$Sepal.Length)
mean.sepal <- iris$Sepal.Length %>% mean()


x <- c(0.109, 0.359, 0.63, 0.996, 0.515, 0.142, 0.017, 0.829, 0.907)
round(exp(diff(log(x))), 1)
x %>% log() %>%
  diff() %>%
  exp() %>%
  round(1)
```

## dplyr, pipe operator

``````{r, echo=T}



# ##########################################################
# ####  dplyr, pipe operator
# ##########################################################
# 
data(starwars)
#
## Look at data: what do rows represent? how about columns?
starwars
# 
# ## filter(): filter data frame on row
filter(starwars, species=="Droid")
starwars[which(starwars$species=="Droid"),] ## base R equivalent
starwars %>% filter(species=="Droid")     ## equivalent with a pipe
starwars %>% filter(., species=="Droid")     ## equivalent with a pipe


starwars %>% filter(species=="Droid" & eye_color=="red")
new_df <- starwars %>% 
  filter(species=="Droid", eye_color=="red") ## 'and' conditions can be separated with comma

```


```{r, echo=T, eval=F}


# ## select(): filter data frame on column
starwars %>% select(mass, homeworld)
starwars %>% select(-hair_color)   ## We can remove rows using a '-'

starwars %>%
  filter(species=="Droid") %>%
  select(name, mass)   ## Filter and select can be combined using a pipe %>%

# ## mutate(): add new column to data frame
# ## Note: we generally avoid overwriting data in a data frame, better to create a new variable
starwars %>%
  mutate(height_in_inches=0.393701 * height) %>%
  View()

# ## group_by(): establish a grouping for downstream operations, remove with ungroup()
# ## tally(): count number of observations per grouping
starwars %>%
  group_by(species) %>%
  tally()

# ## summarize(): perform summary statistic on a column. Fun fact: this also exists as summarise()
starwars %>% summarize(mean_mass = mean(mass, na.rm=TRUE))

starwars %>%
  group_by(species) %>%
  summarize(mean_mass = mean(mass, na.rm=TRUE))

# ## arrange(): arrange a column

starwars %>% arrange(desc(mass))
starwars %>% arrange(mass)

# ## unique, na.omit()
starwars %>% unique()
starwars %>% na.omit()

# #-------------------------------------------------------------------------
# # An in-class example:
# 
# 
# ##########################################################
# ####  If there's time: a quick start on ggplot2
# ##########################################################
# 
# ## ggplot2: graphics package that implements a Grammar of Graphics (operates on dataframes)
# ## Explicit differentiation between data and a representation of the data
# 
# ## ggplot2 grammar:
# ## data = dataframe being plotted
# ## geometrics = the geometric shape that represents the data (point, boxplot, histogram, violin, bar, etc)
# ## aesthetics = aesthetics of the geometric object (color, size, shape, etc)
# 
# ## What are the differences between the ggplot2 and base R versions of this scatterplot?
ggplot(starwars, aes(x=mass, y=height)) +
  geom_point()
plot(starwars$mass, starwars$height) ## base R version

# ## NOTE: did you see that we are using '+' here for ggplot2 and not the pipe '%>%'?

## These are all the same
ggplot() +
  geom_point(data=starwars, color="red", 
             aes(x=mass, y=height))

ggplot() +
  geom_point(data=starwars, color="red",
             aes(x=mass, y=height))

ggplot(data=starwars) +
  geom_point(color="red", aes(x=mass, y=height))

ggplot(starwars, aes(x=mass, y=height)) +
  geom_point(color="red")

## If we place color inside of the aesthetics, it maps it to the data
ggplot(starwars, aes(x=mass, y=height, color=gender)) +
  geom_point()

new_df <- starwars %>% filter(mass < 500)
ggplot(new_df, 
       aes(x=mass, y=height, color=gender)) +
  geom_point()


# ## We can change shape aesthetics too
ggplot(starwars, aes(x=mass, y=height, color=gender, shape=gender)) +
  geom_point()

# ## Aesthetics can be place inside the relevant geom: the following two are equivalent
ggplot(starwars, aes(x=mass, y=height, color=gender, shape=gender)) +
  geom_point()
ggplot(starwars) +
  geom_point(aes(x=mass, y=height, color=gender, shape=gender))

# ## Aesthetics are only for mapping:
ggplot(starwars) +
  geom_point(aes(x=mass, y=height, color="blue"))  ## This doesn't work
ggplot(starwars) +
  geom_point(aes(x=mass, y=height), color="blue")  ## This does work

# ## Can use multiple geoms, combine with dplyr functions
ggplot(filter(starwars, mass < 500), aes(x=mass, y=height)) +
  geom_point()  +
  geom_line()

# ## Other geoms: histograms, boxplots, density plots, violin plots, barplots
ggplot(starwars) +
  geom_histogram(aes(x=height), fill="orange", color="brown")

ggplot(starwars) +
  geom_boxplot(aes(x="", y=log(mass)))

ggplot(starwars) +
  geom_boxplot(aes(x=gender, y=log(mass), fill=gender))

ggplot(starwars) +
  geom_violin(aes(x=gender, y=log(mass), fill=gender))

ggplot(starwars) +
  geom_bar(aes(x=gender, fill=gender))

ggplot(starwars) +
  geom_density(aes(x=log(mass), fill=gender))
ggplot(starwars) +
  geom_density(aes(x=log(mass), fill=gender), alpha=0.9)  ## Can add transparency to object

ggplot(starwars) +
  geom_density(aes(x=log(mass), 
      fill=factor(gender, levels=c("hermaphrodite", "none",
           "NA", "male", "female"))), alpha=0.9)  



# ## Some additional options
ggplot(starwars) +
  geom_histogram(aes(x=height), fill="orange", color="brown") +
  xlab("Height") +
  ylab("Count") +
  ggtitle("Histogram of Star Wars character heights")

g1 <- ggplot(starwars) +
  geom_histogram(aes(x=height), fill="orange", color="brown") 

# ## Change gray background theme: 
# ## http://ggplot2.tidyverse.org/reference/ggtheme.html
# ## https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html
ggplot(starwars) +
  geom_histogram(aes(x=height), fill="orange", color="brown") +
  xlab("Height") +
  ylab("Count") +
  ggtitle("Histogram of Star Wars character heights") +
  theme_minimal()




## Next time: changing the order of plots, changing colors: scale_FILLCOLOR_brewer, 
## scale_color_gradient(), geom_smooth, more advanced dplyr
## http://ggplot2.tidyverse.org/reference/#scales

```
